Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-06-17 paper.tex
minor corrections text

@@

 \C{Also note that the flux upper limits that we compare to are in
   reality a function of spectral indices $\alpha$. However, in the
   relevant energy range $0.1-100$~GeV these spectral indices are
   approximately $1.5 < \alpha < 3.0$, where the change in photon flux
   upper limits is less than 50\%, with the Fermi-LAT being more
   sensitive to the hard spectrum \cite{2010ApJ...717L..71A}. For
   simplicity of the analysis in our paper we adopt the upper limits
   calculated for a spectral index $\alpha=2$.  For consistency we
   estimate the change in the differential upper limits for each
   gamma-ray model and energy range using:
\begin{eqnarray}
\frac{\dd F_\gamma}{\dd \eg} &=& K\,\eg^{-\alpha}\,,\quad F_\gamma = 
K\,\eg^{1-\alpha}\,,\nonumber\\
<\eg \,F_\gamma> &=& K\,\frac{\sum \dd \eg^{2-\alpha}}{\sum \dd \eg}
= K\,\frac{\sum \dd \eg^{2-\alpha}}{E_{\gamma,2}-E_{\gamma,1}}\,,\nonumber
\end{eqnarray}
where $E_{\gamma,1}$ and $E_{\gamma,2}$ are the upper and lower energy
limits, respectively. In the 1-10~GeV energy range we find that the
upper limits for $($LP,BM,CR$)$ change by a factor
$(1.10,\,0.71,\,1.15)$ where we have used the spectral indices
$(2.20,\,1.22,\,2.24)$, respectively. See Fig.~\ref{fig:clu_comp} for
additional spectral indices.}

@@

What is the figure of merit for selecting the most promising cluster
targets for indirect DM searches? To this end, we employ the
luminosity-to-mass scaling relations. The gamma-ray luminosity from
the smooth density profile is given by \cite{2009PhRvL.103r1302P}
\begin{equation}
L_{\gamma,\sm} \propto \int \dd V \rho(r)^2 \propto \frac{M_{200}\,c^3}
{\left[\log\left(1+c\right)-c/(1+c)\right]^2} \propto \mvir^{0.83}\,,
\end{equation}
\del{and the total DM luminosity that includes boost factors for the LP and
the BM models are given by}
%\begin{eqnarray}
%\label{eq:LP_scaling}
%L_{\gamma,\rmn{LP}} &=& L_{\gamma,\rmn{smLP}} \B_\rmn{LP} \propto \frac{\mvir^{0.73}}{D_\rmn{lum}^2},\\
%\label{eq:BM_scaling}
%L_{\gamma,\rmn{BM}} &=& L_{\gamma,\rmn{smBM}} \B_\rmn{BM} \propto  \frac{\mvir^{1.06}}{D_\rmn{lum}^2},\\
%\B_\rmn{LP}&=&\B_\sfe \B_\sub \propto \mvir^{-1/3}\mvir^{0.23} \propto
% \mvir^{-0.1}\,,\\
%\B_\rmn{BM}&=&\B_\sub \propto \mvir^{0.23}\,.
%\end{eqnarray}
\del{Hence, increasing the cluster mass in the BM model by a factor two
more than doubles the luminosity, but increasing the luminosity
distance, $D_\rmn{lum}$, by a factor two decrease the flux by a factor
four. We note that the IC from upscattered SD photons scales slightly
softer with mass, hence gives rise to a larger fraction of gamma-rays
in low mass clusters compared to the total gamma-ray flux.}
\C{and the total DM luminosity that includes boost factors for the LP and
the BM models is given by
\begin{eqnarray}
\label{eq:DM_scaling}
L_{\gamma} &=& L_{\gamma,\rmn{sm}} \B_\rmn{sub} \propto \frac{\mvir^{1.06}}{D_\rmn{lum}^2},\\
\rmn{where} & &\B_\sub \propto \mvir^{0.23}\,.
\end{eqnarray}
Hence, increasing the cluster mass by a factor two more than doubles
the DM luminosity, but increasing the luminosity distance,
$D_\rmn{lum}$, by a factor two decrease the flux by a factor four. We
note that the IC from upscattered SD photons scales slightly softer
with mass, hence gives rise to a larger fraction of gamma-rays in low
mass clusters compared to the total gamma-ray flux. Also note that we
have not included the SFE for the LP model in
Eq.~(\ref{eq:DM_scaling}) since it is a constant when substructures
are included. If only a small fraction of the DM resides in the
subhalos, the scaling of the LP DM becomes slightly steeper due to the
mass dependence inferred from the velocity dispersion:
\begin{eqnarray}
L_{\gamma,LPnosub} &=& L_{\gamma,\rmn{sm}} \B_\rmn{sfe} 
\propto \frac{\mvir^{0.5}}{D_\rmn{lum}^2},\\
\rmn{where} &\quad&\B_\rmn{sfe} \propto \mvir^{-1/3}.
\end{eqnarray}}

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Fornax.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Coma.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Virgo.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Perseus.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\caption{\it Comparing the gamma-ray emission from different
  clusters. We show the differential flux of clusters: Fornax (upper
  left), Coma (upper right), Virgo (lower left), and Perseus (lower
  right). We show CR induced emission (red solid), a leptophilic (LP)
  model that includes both final state radiation and IC upscattered
  CMB, dust and starlight (green dash-dotted), and the benchmark $\Kp$
  model (BM) that includes continuum emission, and IC upscattered CMB,
  dust and starlight (blue dashed). The arrows show the spatially
  extended differential upper limits set by Fermi-LAT in the energy
  ranges $0.2-1$~GeV, $1-10$~GeV, and $10-100$~GeV from left to
  right. We show the individual components for the BM model and LP
  model for Virgo and Perseus, respectively. The flux from the
  clusters is integrated out to $\rvir$ using a point spread function
  of $0.1\deg$ and includes the boost from substructures. We find that
  the lower GeV energy regime is most constraining due to the peak
  sensitivity of Fermi at these energies and use this regime to get
  upper limits on boost factors.}
 \label{fig:clu_comp}
\end{minipage}
\end{figure*}

@@

\C{In Fig.~\ref{fig15}, we show the extended Fermi-LAT flux upper
  limits that assume King profiles that trace the X-ray emitting gas
  compared to the flux predictions by an analytical CR model
  \cite{2010MNRAS.409..449P} for CR induced gamma-rays.} For
consistency, we use beta-model fits to ROSAT data of X-ray bright
clusters for our gamma-ray flux estimates as an input
\cite{2002ApJ...567..716R}). We derive the central gas densities for
each cluster by following the approach outlined in
\cite{1999ApJ...517..627M} which assumes that the ICM is
isothermal. The central gas density is found by combining X-ray
temperature with X-ray luminosity, X-ray detection radius, density
core radius and density slope for each cluster. These predictions are
superior to those that use luminosity-mass scaling relation to infer
the expected gamma-ray flux---a method used by reference
\cite{2010ApJ...717L..71A} to compare to Fermi limits.  In this
analytic model, we employ the fact that CRs accelerated at structure
formation shocks and transported over cosmic time. This leads to a CR
distribution with a universal spectral form and an almost universal
spatial distribution. Depending on whether or not we account for the
bias of ``artificial galaxies'' in cosmological simulations, we derive
an optimistic or conservative limit of the expected gamma-ray emission
from decaying pions (bracketing our ignorance of the contribution of
cluster galaxies to the overall gamma-ray flux from a cluster). Note
that both models assume the maximally allowed acceleration efficiency
for strong shocks that transfers 50\% of the shock-dissipated energy
(kinetic energy corrected for adiabatic compressional heating at the
shock) to CRs. These efficiencies drop quickly for weaker shocks
\cite{2007A&A...473...41E} which dominate the gravitational energy
dissipation inside galaxy clusters \cite{2006MNRAS.367..113P}. These
efficiencies will have to be corrected down if this value is not
realized at structure formation shocks. Secondly, these simulations
neglect active CR transport such as streaming and diffusion relative
to the gas, i.e. we assume that advective transport is the dominant
one and CRs are tightly coupled to the gas via flux-frozen magnetic
fields tangled on sufficiently small scales. Evidence from the
statistics of radio halos suggests that this might not be the case for
relaxed clusters, arguing that CR streaming in those objects could be
responsible for a decrease in the central CR number density. This
effect would also reduce the expected gamma-ray flux by up to a factor
of five \cite{2011A&A...527A..99E}. We note that the tightest Fermi
limits are obtained in the 1-10 GeV regime due to the highest
sensitivity of Fermi-LAT there. While the effective area of Fermi-LAT
increases up to these energies, the typical photon spectra decrease as
a function of energy: combining those effects selects this energy
range to be most sensitive.  In this energy band, the Fermi limits
come close to the flux predictions of Virgo, Coma and Norma. Using
integrated flux limits from Fermi, we arrive at very similar results.
We emphasize that by using our analytic modeling, the slight
discrepancy of the scaling relation-based prediction with the Fermi
limit on Norma \cite{2010ApJ...717L..71A} is resolved: Norma simply
seems to be less bright than an average cluster of the same mass as
Norma. The insensitive limit on Perseus \C{is due to the bright
  central source NGC1275 with an AGN in the center
  \cite{2010ATel.2916....1M} making it hard to determine a comparable
  limit as obtained for the other clusters. In addition, there are
  recent gamma-ray observations of the head-tail radio galaxy IC310 in
  the cluster outskirts
  \cite{2010ApJ...723L.207A,2010A&A...519L...6N}.}

\C{A quantity that is of great scientific interest is the CR pressure
  ($P_\CR$) relative to the thermal pressure ($P_\rmn{th}$), $X_\CR =
  P_\CR/P_\rmn{th}$, as it directly bias galaxy cluster observables
  and the resulting cosmological constraints, such as the hydrostatic
  mass bias or the Sunyaev-Zel’dovich signal measured by Planck, South
  Pole Telescope, or Atacama Cosmology Telescope. In the following we
  estimate the mass scaling of the relative CR pressure using the CR
  formalism outlined in
  \cite{2007A&A...473...41E,2010MNRAS.409..449P}.  First we calculate
  the CR distribution function $f = C\,p^{-\alpha}$, where $C$ is the
  normalization of each power-law, $p=P/m_\rmn{p} c$ is the normalized
  CR proton momentum, and $\alpha$ is the spectral index. The CR
  pressure $P_\CR \propto C=\tilde{C}/\rho_\rmn{gas}$, where
  $\tilde{C}\propto M^{0.44}$ for clusters with a mass $\mvir\gtrsim
  10^{14}\,\msun$. We determine the mass scaling of the gas density
  through its relation to the enclosed gas fraction $f_\rmn{gas}$, and
  find that $\rho_\rmn{gas}\propto f_\rmn{gas} \propto M^{0.135}$
  \cite{2009ApJ...693.1142S}. The thermal pressure is derived from the
  temperature in virial equilibrium, and scales as $P_\rmn{th}\propto
  k_\B T \propto G M/R \propto M^{0.67}$. Finally we can derive the
  mass scaling of the volume weighted relative CR pressure from
\begin{equation}
 <X_\CR> = \frac{<P_\CR>}{<P_\rmn{th}>} \propto \frac{<\rho_\rmn{gas}\,
   \tilde{C}>}{<k_\B T>} \propto M^{-0.06}\,,
\end{equation}
which is roughly a constant, i.e. almost independent on mass. Hence we
adopt a constant $<X_\CR>\approx 0.02$ derived from a representative
simulated cluster \cite{2008MNRAS.385.1211P}.  Since the CRs that
build up the bulk of the CR pressure also give rise to gamma-ray
emission through the production of decaying pions, we use gamma-ray
upper limits from Fermi-LAT to constrain $X_\CR$. In
Fig.~\ref{fig:XCR} we show the volume weighted relative CR pressure
for 32 clusters, for which LAT upper limits that assume King profiles
are derived. We find that the most constraining clusters are the Norma
and Coma clusters of $<X_\CR> < 0.07$.}

@@

We substantially improve the modelling of the expected gamma-ray
signal from pion-decay resulting from hadronic CR interactions with
gas protons over previous work \citep{2010ApJ...717L..71A}. We employ
the analytic model of the CR spectrum and spatial distribution,
recently developed by \citet{2010MNRAS.409..449P} that is based on
cosmological hydrodynamical simulations of galaxy clusters that
self-consistently follow the evolution of CRs in cosmic structure
formation. Adopting the model for all clusters in the HIFLUGCS sample
of flux-limited X-ray clusters, we compare the expected gamma-ray
emission to that inferred from using $L_\gamma$-cluster mass scaling
relation. We note that the predictions of our ``optimistic'' model are
perfectly in agreement with Fermi upper limits for all clusters.  The
hint of a tension that the scaling relation modeling showed for Norma
\citep{2010ApJ...717L..71A} could not be confirmed due to the lower
gas density of this cluster relative to the average.  As expected, the
analytical CR model accounts for the ``scatter'' to the scaling
relation and biases the gamma-ray flux high for prominent cool core
clusters by up to a factor of 6 relative to the expectation from the
scaling relation. This is due to the high target gas densities and
associated CR densities in those cool cores that shape the spatial
emission characteristic to be very similar to that observed in thermal
X-rays. We caution that these predictions only apply for the case of
negligible active CR transport such as CR streaming and diffusion
relative to the gas. Gamma-ray fluxes might be considerably smaller in
those cluster that do not show an extended diffuse radio(-mini) halo
that might point to a centrally concentrated CR distribution
\citep{2011A&A...527A..99E}. The CR spectrum $E^2 dN/dE$ shows the
characteristic pion bump at energies around 1~GeV. It is expected to
dominate the DM annihilation signal of BM models for all
clusters/groups but Fornax. \C{We conclude that due to the large
  spatial extension of the brightest clusters it is not possible for
  Fermi to rule out the LP model nor to constrain the BM DM models
  without improved modelling. However, we can put tight constrints on
  the boost from substructures and Sommerfeld enhancement using the LP
  model. Hence, clusters are a nice complement to direct and
  accelerator searches for the nature of DM.}\del{We conclude that it
  will be possible to rule out the LP model with six years of Fermi
  data and start to constrain some BM DM models. Hence this should
  nicely complement direct and accelerator searches for the nature of
  DM.}

>>>>

%\voffset-.6in 
\voffset.3in

@@

 \C{Also note that the flux upper limits that we compare to are in
   reality a function of spectral indices $\alpha$. However, in the
   relevant energy range $0.1-100$~GeV these spectral indices are
   approximately $1.5 < \alpha < 3.0$, where the change in photon flux
   upper limits is less than 50\%, with the Fermi-LAT being more
   sensitive to the hard spectrum \cite{2010ApJ...717L..71A}. For
   simplicity of the analysis in our paper we adopt the upper limits
   calculated for a spectral index $\alpha=2$.  For consistency we
   estimate the change in the differential upper limits for each
   gamma-ray model and energy range using:
\begin{eqnarray}
\frac{\dd F_\gamma}{\dd \eg} &=& K\,\eg^{-\alpha}\,,\quad F_\gamma = 
K\,\eg^{1-\alpha}\,,\nonumber\\
<\eg \,F_\gamma> &=& K\,\frac{\sum \dd \eg^{2-\alpha}}{\sum \dd \eg}
= K\,\frac{\sum \dd \eg^{2-\alpha}}{E_{\gamma,2}-E_{\gamma,1}}\,,\nonumber\\
\label{eq:spec_ind_UL}
\end{eqnarray}
where $E_{\gamma,1}$ and $E_{\gamma,2}$ are the upper and lower energy
limits, respectively. In the 1-10~GeV energy range we find that the
upper limits for $($LP,BM,CR$)$ change by a factor
$(1.10,\,0.71,\,1.15)$ where we have used the spectral indices
$(2.20,\,1.22,\,2.24)$, respectively. See
Tab.~\ref{tab:spectral_index} for additional spectral indices.}

@@

What is the figure of merit for selecting the most promising cluster
targets for indirect DM searches? To this end, we employ the
luminosity-to-mass scaling relations. The gamma-ray luminosity from
the smooth density profile is given by \cite{2009PhRvL.103r1302P}
\begin{equation}
L_{\gamma,\sm} \propto \int \dd V \rho(r)^2 \propto \frac{M_{200}\,c^3}
{\left[\log\left(1+c\right)-c/(1+c)\right]^2} \propto \mvir^{0.83}\,,
\end{equation}
\del{and the total DM luminosity that includes boost factors for the LP and
the BM models are given by}
%\begin{eqnarray}
%\label{eq:LP_scaling}
%L_{\gamma,\rmn{LP}} &=& L_{\gamma,\rmn{smLP}} \B_\rmn{LP} \propto \frac{\mvir^{0.73}}{D_\rmn{lum}^2},\\
%\label{eq:BM_scaling}
%L_{\gamma,\rmn{BM}} &=& L_{\gamma,\rmn{smBM}} \B_\rmn{BM} \propto  \frac{\mvir^{1.06}}{D_\rmn{lum}^2},\\
%\B_\rmn{LP}&=&\B_\sfe \B_\sub \propto \mvir^{-1/3}\mvir^{0.23} \propto
% \mvir^{-0.1}\,,\\
%\B_\rmn{BM}&=&\B_\sub \propto \mvir^{0.23}\,.
%\end{eqnarray}
\del{Hence, increasing the cluster mass in the BM model by a factor two
more than doubles the luminosity, but increasing the luminosity
distance, $D_\rmn{lum}$, by a factor two decrease the flux by a factor
four. We note that the IC from upscattered SD photons scales slightly
softer with mass, hence gives rise to a larger fraction of gamma-rays
in low mass clusters compared to the total gamma-ray flux.}
\C{and the total DM luminosity that includes boost factors for the LP and
the BM models is given by
\begin{eqnarray}
\label{eq:DM_scaling}
L_{\gamma} &=& L_{\gamma,\rmn{sm}} \B_\rmn{sub} \propto \frac{\mvir^{1.06}}{D_\rmn{lum}^2},\\
\rmn{where} & &\B_\sub \propto \mvir^{0.23}\,.
\end{eqnarray}
Hence, increasing the cluster mass by a factor two more than doubles
the DM luminosity, but increasing the luminosity distance,
$D_\rmn{lum}$, by a factor two decrease the flux by a factor four. We
note that the IC from upscattered SD photons scales slightly softer
with mass, hence gives rise to a larger fraction of gamma-rays in low
mass clusters compared to the total gamma-ray flux. Also note that we
have not included the SFE for the LP model in
Eq.~(\ref{eq:DM_scaling}) since it saturates to a constant value in
the substructures. However, if only a small fraction of the DM resides
in the subhalos, the scaling of the LP DM becomes slightly steeper due
to the mass dependence inferred from the velocity dispersion:
\begin{eqnarray}
L_{\gamma,LPnosub} &=& L_{\gamma,\rmn{sm}} \B_\rmn{sfe} 
\propto \frac{\mvir^{0.5}}{D_\rmn{lum}^2},\\
\rmn{where} &\quad&\B_\rmn{sfe} \propto \mvir^{-1/3}.
\end{eqnarray}}

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Fornax.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Coma.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Virgo.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Perseus.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\caption{\it Comparing the gamma-ray emission from different
  clusters. We show the differential flux of clusters \C{in the upper
    panel of each figure}: Fornax (upper left), Coma (upper right),
  Virgo (lower left), and Perseus (lower right). We show CR induced
  emission (red solid), a leptophilic (LP) model that includes both
  final state radiation and IC upscattered CMB, dust and starlight
  (green dash-dotted), and the benchmark $\Kp$ model (BM) that
  includes continuum emission, and IC upscattered CMB, dust and
  starlight (blue dashed). The arrows show the spatially extended
  differential upper limits set by Fermi-LAT in the energy ranges
  $0.2-1$~GeV, $1-10$~GeV, and $10-100$~GeV from left to right. We
  show the individual components for the BM model and LP model for
  Virgo and Perseus, respectively. \C{In the lower panel of each
    figure we show differential spectral index where $\dd F_\gamma/\dd
    E_\gamma \sim E_\gamma^{-\alpha_\gamma}$.} The flux from the
  clusters is integrated out to $\rvir$ using a point spread function
  of $0.1\deg$ and includes the boost from substructures. We find that
  the lower GeV energy regime is most constraining due to the peak
  sensitivity of Fermi at these energies and use this regime to get
  upper limits on boost factors.}
 \label{fig:clu_comp}
\end{minipage}
\end{figure*}

\C{For consistency we provide in addition the differential spectral
  indices $\alpha_\gamma=-\frac{\dd \log(\dd F/\dd E)}{\dd \log(E)}$
  for the different emission models in the lower panels of
  Fig.~\ref{fig:clu_comp}. We find very similar spectral indices for
  different clusters when the boost from substructures is substantial,
  while for models without substructures the relative contribution
  from the upscattered dust and starlight breaks the spectral
  universality for the DM models. In the $1~\gev-1~\tev$ energy regime
  the CR spectral index is approximately constant $\sim (2.1-2.3)$,
  while the spectral index for the LP model has a large spread and
  varies between $1.2-4.0$, hence implying that gamma-ray upper limits
  are more sensitive to the specific energy regime for these
  models. Similarly, the index for the BM $\Kp$ model is about 1.0 at
  1~GeV and increases monotonically for higher energies, which also
  suggests that the flux upper limits contrasted to this model are
  sensitive to the specific energy range, and where a small energy
  range would be preferred. In experiments is the spectral index often
  fixed over an energy range, hence we show in
  Tab.~\ref{tab:spectral_index} the indices for our emission models in
  four different energy bands. We find that these spectral indices
  have a much smaller spread, hence reducing the importance of the
  specific energy regimes. For further discussion about the impact of
  different spectral indices on the gamma-ray upper limits see
  Eq.~(\ref{sect:spectral}) and related text.}

\begin{table}
\begin{tabular}{ccccc}
\hline\hline
      Model & $\alpha_{100\,\rmn{MeV}}^{1\,\rmn{GeV}}$  & 
              $\alpha_{1\,\rmn{GeV}}^{10\,\rmn{GeV}}$   & 
              $\alpha_{10\,\rmn{GeV}}^{100\,\rmn{GeV}}$ &
              $\alpha_{100\,\rmn{GeV}}^{1\,\rmn{TeV}}$  \\
\hline
CR-$\pi^0$      & 1.44 & 2.24 & 2.30 & 2.26 \\
DM-LP Sum       & 1.74 & 2.20 & 3.06 & 1.30 \\
DM-BM-$\Kp$ Sum  & 1.28 & 1.22 & 2.23 & -    \\
\hline\hline
\end{tabular}
 \caption{\C{Gamma-ray spectral indices $\alpha_{E_1}^{E_2}$ between
     energies $E_1$ and $E_2$. We show the spectral index for three
     different emission models where the boost from substructures is
     included; the CR induced emission, the leptophilic (LP) model
     that includes both final state radiation and IC upscattered CMB,
     dust and starlight, and the benchmark (BM) $\Kp$ model that
     includes continuum emission, and IC upscattered CMB, dust and
     starlight. The indices are derived for the Fornax cluster,
     although the variance between clusters is very small when the
     enhancement from substructures is included which suppresses the
     relative contribution from the upscattered dust and starlight
     component that is cluster dependent. \label{tab:spectral_index}}}
\end{table}

@@

\C{In Fig.~\ref{fig15}, we show the extended Fermi-LAT flux upper
  limits that assume King profiles that trace the X-ray emitting gas
  compared to the flux predictions by an analytical CR model
  \cite{2010MNRAS.409..449P} for CR induced gamma-rays.} For
consistency, we use beta-model fits to ROSAT data of X-ray bright
clusters for our gamma-ray flux estimates as an input
\cite{2002ApJ...567..716R}). We derive the central gas densities for
each cluster by following the approach outlined in
\cite{1999ApJ...517..627M} which assumes that the ICM is
isothermal. The central gas density is found by combining X-ray
temperature with X-ray luminosity, X-ray detection radius, density
core radius and density slope for each cluster. These predictions are
superior to those that use luminosity-mass scaling relation to infer
the expected gamma-ray flux---a method used by reference
\cite{2010ApJ...717L..71A} to compare to Fermi limits.  In this
analytic model, we employ the fact that CRs accelerated at structure
formation shocks and transported over cosmic time. This leads to a CR
distribution with a universal spectral form and an almost universal
spatial distribution. Depending on whether or not we account for the
bias of ``artificial galaxies'' in cosmological simulations, we derive
an optimistic or conservative limit of the expected gamma-ray emission
from decaying pions (bracketing our ignorance of the contribution of
cluster galaxies to the overall gamma-ray flux from a cluster). Note
that both models assume the maximally allowed acceleration efficiency
for strong shocks that transfers 50\% of the shock-dissipated energy
(kinetic energy corrected for adiabatic compressional heating at the
shock) to CRs. These efficiencies drop quickly for weaker shocks
\cite{2007A&A...473...41E} which dominate the gravitational energy
dissipation inside galaxy clusters \cite{2006MNRAS.367..113P}. These
efficiencies will have to be corrected down if this value is not
realized at structure formation shocks. Secondly, these simulations
neglect active CR transport such as streaming and diffusion relative
to the gas, i.e. we assume that advective transport is the dominant
one and CRs are tightly coupled to the gas via flux-frozen magnetic
fields tangled on sufficiently small scales. Evidence from the
statistics of radio halos suggests that this might not be the case for
relaxed clusters, arguing that CR streaming in those objects could be
responsible for a decrease in the central CR number density. This
effect would also reduce the expected gamma-ray flux by up to a factor
of five \cite{2011A&A...527A..99E}. We note that the tightest Fermi
limits are obtained in the 1-10 GeV regime due to the highest
sensitivity of Fermi-LAT there. While the effective area of LAT
increases up to these energies, the typical photon spectra decrease as
a function of energy: combining those effects selects this energy
range to be most sensitive.  In this energy band, the Fermi limits
come close to the flux predictions of Virgo, Coma and Norma. Using
integrated flux limits from Fermi, we arrive at very similar results.
We emphasize that by using our analytic modeling, the slight
discrepancy of the scaling relation-based prediction with the Fermi
limit on Norma \cite{2010ApJ...717L..71A} is resolved: Norma simply
seems to be less bright than an average cluster of the same mass as
Norma. The insensitive limit on Perseus \C{is due to the bright
  central source NGC1275 with an AGN in the center
  \cite{2010ATel.2916....1M} making it hard to determine a comparable
  limit as obtained for the other clusters. In addition, there are
  recent gamma-ray observations of the head-tail radio galaxy IC310 in
  the cluster outskirts
  \cite{2010ApJ...723L.207A,2010A&A...519L...6N}.}

\C{A quantity that is of great scientific interest is the CR pressure
  ($P_\CR$) relative to the thermal pressure ($P_\rmn{th}$) as it
  directly bias galaxy cluster observables and the resulting
  cosmological constraints, such as the hydrostatic mass bias or the
  Sunyaev-Zel’dovich signal measured by Planck, South Pole Telescope,
  or Atacama Cosmology Telescope. In the following we estimate the
  mass scaling of the relative CR pressure, $X_\CR =
  P_\CR/P_\rmn{th}$, using the CR formalism outlined in
  \cite{2007A&A...473...41E,2010MNRAS.409..449P}.  First we calculate
  the CR distribution function $f = C\,p^{-\alpha}$, where $C$ is the
  normalization of each power-law, $p=P/m_\rmn{p} c$ is the normalized
  CR proton momentum, and $\alpha$ is the spectral index. The CR
  pressure $P_\CR \propto C=\tilde{C}/\rho_\rmn{gas}$, where
  $\tilde{C}\propto M^{0.44}$ for clusters with a mass $\mvir\gtrsim
  10^{14}\,\msun$ \cite{2010MNRAS.409..449P}. We determine the
  mass scaling of the gas density through its relation to the enclosed
  gas fraction $f_\rmn{gas}$, and find that $\rho_\rmn{gas}\propto
  f_\rmn{gas} \propto M^{0.135}$ \cite{2009ApJ...693.1142S}. The
  thermal pressure is derived from the temperature in virial
  equilibrium, and scales as $P_\rmn{th}\propto k_\B T \propto G M/R
  \propto M^{0.67}$. Finally we can derive the mass scaling of the
  volume weighted relative CR pressure from
\begin{equation}
 <X_\CR> = \frac{<P_\CR>}{<P_\rmn{th}>} \propto \frac{<\rho_\rmn{gas}\,
   \tilde{C}>}{<k_\B T>} \propto M^{-0.06}\,,
\end{equation}
which is roughly a constant, i.e. almost independent on mass. Hence we
adopt a constant $<X_\CR>\approx 0.02$ derived from a representative
simulated cluster \cite{2008MNRAS.385.1211P}. Since the CRs that build
up the bulk of the CR pressure also give rise to gamma-ray emission
through the production of decaying pions, we use gamma-ray upper
limits from Fermi-LAT to constrain the volume weighted $X_\CR$. In
Fig.~\ref{fig:XCR} we show the volume weighted relative CR pressure
for 32 clusters, for which LAT upper limits that assume King profiles
are derived. For most of the clusters in the sample we can constrain
$X_\CR < (0.1-0.2)$ using the extended gamma-ray upper limits, where
the most constraining clusters are Norma and Coma with $X_\CR <
0.07$. Most of the 32 clusters have $X_\CR$ upper limits within a
factor few from the best constraining clusters, which suggests that
stacking of clusters will be able to improve these limits. This might
even be improved further using a scaled stacking with our model as a
prior. In previous work \cite{2010ApJ...717L..71A} where the 1.5 year
Fermi data was used to derive limits on $X_\CR$ they found on average
more constraining limits on the most massive clusters, while we find
better limits on the low mass clusters. This difference is explained
by the different modelling of $P_\CR$; they assumed a mass
independent CR pressure while we assume that $P_\CR\sim
M^{0.44}$. Furthermore, the gamma-ray flux predictions that we use in
this work to constrain $X_\CR$ assumes the true gas density
distribution for each cluster \footnote{Derived from the parameters in
  the extended HIFLUGCS catalogue.} that accounts for
cluster-to-cluster scatter, in contrast to the mass-to-luminosity
scaling relations that was used in previous works.}

@@

We substantially improve the modelling of the expected gamma-ray
signal from pion-decay resulting from hadronic CR interactions with
gas protons over previous work \citep{2010ApJ...717L..71A}. We employ
the analytic model of the CR spectrum and spatial distribution,
recently developed by \citet{2010MNRAS.409..449P} that is based on
cosmological hydrodynamical simulations of galaxy clusters that
self-consistently follow the evolution of CRs in cosmic structure
formation. Adopting the model for all clusters in the HIFLUGCS sample
of flux-limited X-ray clusters, we compare the expected gamma-ray
emission to that inferred from using $L_\gamma$-cluster mass scaling
relation. We note that the predictions of our ``optimistic'' model are
perfectly in agreement with Fermi upper limits for all clusters.  The
hint of a tension that the scaling relation modeling showed for Norma
\citep{2010ApJ...717L..71A} could not be confirmed due to the lower
gas density of this cluster relative to the average.  As expected, the
analytical CR model accounts for the ``scatter'' to the scaling
relation and biases the gamma-ray flux high for prominent cool core
clusters by up to a factor of 6 relative to the expectation from the
scaling relation. This is due to the high target gas densities and
associated CR densities in those cool cores that shape the spatial
emission characteristic to be very similar to that observed in thermal
X-rays. We caution that these predictions only apply for the case of
negligible active CR transport such as CR streaming and diffusion
relative to the gas. Gamma-ray fluxes might be considerably smaller in
those cluster that do not show an extended diffuse radio(-mini) halo
that might point to a centrally concentrated CR distribution
\citep{2011A&A...527A..99E}. The CR spectrum $E^2 dN/dE$ shows the
characteristic pion bump at energies around 1~GeV. It is expected to
dominate the DM annihilation signal of BM models for all
clusters/groups but Fornax. \C{We conclude that due to the large
  spatial extension of the brightest clusters it is not possible for
  Fermi to rule out the LP model nor to constrain the BM DM models
  without improved modelling. However, we can put tight constraints on
  the boost from substructures and Sommerfeld enhancement using the LP
  model. Hence, clusters are a nice complement to direct and
  accelerator searches for the nature of DM.}\del{We conclude that it
  will be possible to rule out the LP model with six years of Fermi
  data and start to constrain some BM DM models. Hence this should
  nicely complement direct and accelerator searches for the nature of
  DM.}
